preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2020-09-28 Manuscript/manuscript.Rmd
grammar edits to methods

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://dozmorovlab.github.io/preciseTAD/ and Bioconductor (submitted).

@@

TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools were used for training. The autosomal genomic coordinates in the GRCh37/hg19 human genome assembly were considered. Arrowhead-defined TAD boundaries were called from the GM12878 and K562 cell line Hi-C data (MAPQ>0, 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions) using the default parameters (Additional file 1: Arrowhead Script). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Additional file 2: Table S1<!--Supplementary Table 1-->). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin states (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Additional file 4: Table S3<!--Supplementary Table 2-->).

@@

Cell line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the average signal strength of the corresponding annotation (*Peak Signal Strength*), the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin states lack signal values (Additional file 5: Figure S1<!--Supplementary Figure 1-->).

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and resampling techniques. Following recommendations to evaluate the model on unseen data [@schreiber2019pitfall], a *holdout chromosome* technique was used for estimating model's performance. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector, **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$, and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 0.1% of the rows in the data. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities is obtained from the confusion matrix created by validating the model on the test data. The process was repeated for each $i^{th}$ holdout chromosome, and performances were aggregated using mean and standard deviation.

@@

We used the same holdout chromosome strategy to evaluate a model trained in one cell line on unseen data from another cell line [@schreiber2019pitfall]. Given two cell lines, GM12878 and K562, we first evaluated the performance of cell line-specific models. That is, models trained on cell line-specific data from $n-1$ chromosomes were evaluated on the $i^{th}$ holdout chromosome data from the same cell line. Second, we evaluated models trained on cell line-specific data from $n-1$ chromosomes using the $i^{th}$ holdout chromosome data from a different cell line. That is, models trained using K562 cell line-specific data were evaluated on unseen chromosome data from GM12878 cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivities and specificities at different cutoffs, across each holdout chromosome, and reported the corresponding average area under the curve (AUC).

@@

To investigate whether we could alleviate the resolution limitations of conventional domain calling tools, we developed *preciseTAD*. This algorithm leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution (Additional file 8: Figure S4<!--Supplementary Figure 2-->).

A random forest model $(M)$ was built on the optimal combination of predictor type (predictor space), resampling technique (class balancing correction), and top-ranked annotations (most informative genomic elements) for a set of binned chromosomes $\{k | k \ne i \}$ <!--, to maximally accurately predict boundaries at the _Hi-C resolution level_-->. To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for the chromosome *i*, $A_{n \times p}$, where $n$ is the length of chromosome $i$ and $p$ is the optimal number of annotations. We then evaluate $M$ on the base-level predictor space to extract the probability vector, $\pi_{n_{i}}$, denoting each base's probability of being a boundary. Here, $n_{i}$ represents the length of chromosome *i*. A threshold $t$ specifies the probability at which a base with $\pi_{n_{i}} \ge t$ is designated as a potential boundary (the default $t=1$). Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between boundary-annotated bases, $D$. The minimum and maximum coordinates of each cluster, $k$, of spatially colocalized bases were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* implements partitioning around medoids (PAM) on the distance matrix, $D_{k}$, derived from each cluster. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

@@

$$NE = \dfrac{1}{p} \left[ \Sigma_{s=1}^{p} \left[ \dfrac{1}{b} \Sigma_{k=1}^{b} e_{ks} \times c_{k} \right] \right]$$

@@

where A and B represent genomic regions created by flanked called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both Arrowhead and Peakachu boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The significance level was set to 0.05 for all statistical tests.

>>>>

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://dozmorovlab.github.io/preciseTAD/ and Bioconductor.

@@

TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools were used for training and testing. The autosomal genomic coordinates in the GRCh37/hg19 human genome assembly were considered. Arrowhead-defined TAD boundaries were called from Hi-C data for the GM12878 and K562 cell lines (MAPQ>0, 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions) using the default parameters (Additional file 1: Arrowhead Script). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Additional file 2: Table S1<!--Supplementary Table 1-->). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin states (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Additional file 4: Table S3<!--Supplementary Table 2-->).

@@

Cell line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the average signal strength of the corresponding annotation (*Peak Signal Strength*), the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions  (Additional file 5: Figure S1<!--Supplementary Figure 1-->). Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin states lack signal values.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and resampling techniques. Following recommendations to evaluate the model on unseen data [@schreiber2019pitfall], a *holdout chromosome* technique was used for estimating model performance. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. The number of annotations, $p$, together with the response vector, **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$, and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 0.1% of the rows in the data. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities is obtained from the confusion matrix created by validating the model on the test data. The process was repeated for each $i^{th}$ holdout chromosome, and performances were aggregated using the mean and standard deviation.

@@

We used the same holdout chromosome strategy to evaluate a model trained in one cell line on unseen data from another cell line [@schreiber2019pitfall]. Given two cell lines, GM12878 and K562, we first evaluated the performance of cell line-specific models. That is, models trained on cell line-specific data from $n-1$ chromosomes were evaluated on the $i^{th}$ holdout chromosome data from the same cell line. Second, we evaluated models trained on cell line-specific data from $n-1$ chromosomes using the $i^{th}$ holdout chromosome data from a different cell line. That is, models trained using K562 cell line-specific data were evaluated on unseen chromosome data from the GM12878 cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivities and specificities at different cutoffs, across each holdout chromosome, and reported the corresponding average area under the curve (AUC).

@@

To investigate whether we could alleviate the resolution limitations of conventional domain calling tools, we developed *preciseTAD*. This algorithm leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base-level resolution (Additional file 8: Figure S4<!--Supplementary Figure 2-->).

A random forest model $(M)$ was built on the optimal combination of predictor type (predictor space), resampling technique (class balancing correction), and top-ranked annotations (most informative genomic elements) for a set of binned chromosomes $\{k | k \ne i \}$ <!--, to maximally accurately predict boundaries at the _Hi-C resolution level_-->. To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for the chromosome *i*, $A_{n \times p}$, where $n$ is the length of chromosome $i$ and $p$ is the optimal number of annotations. We then evaluate $M$ on the base-level predictor space to extract the probability vector, $\pi_{n}$, denoting each base's probability of being a boundary. A threshold $t$ specifies the probability at which a base with $\pi_{n} \ge t$ is designated as a potential boundary (the default $t=1$). Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between boundary-annotated bases, $D$, such that $\pi_{n} \ge t$. The minimum and maximum coordinates of each cluster, $k$, of spatially colocalized bases were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* implements partitioning around medoids (PAM) on the distance matrix, $D_{k}$, derived from each cluster. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

@@

$$NE = \dfrac{1}{p} \left[ \Sigma_{s=1}^{p} \left[ \dfrac{1}{b} \Sigma_{k=1}^{b} e_{ks} \right] \right]$$

@@

where A and B represent genomic regions created by flanked called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both Arrowhead and Peakachu boundaries, respectively. All statistical analyses were performed in R (version 4.0.1). The significance level was set to 0.05 for all statistical tests.
